---
title: ABLG methylation analyses using all reference genome comined into one fasta
  file
author: "charles-gumbi"
date: "2025-08-13"
output:
  html_document:
    df_print: paged
---
```{r Loading_packages}
library(dplyr)
library(ggplot2)
library(readr)
library(hrbrthemes)
library(gcookbook)
library(extrafont)
library(tidyverse)
library(hunspell)
```

# Modifications from last analyses
1.Removed T4 and XP-12 phages from the in reference alignment

NEB tech support “We include T4 and XP-12 phages in our control references for hydroxymethylation assessment (a related product). Since you did not spike in T4 or XP-12 to your libraries you can likely ignore any reads that that happen to align to these contigs. If your fin samples do actually contain phages related to XP-12 and T4, you may see more coverage. If you’re interested in the phage genomics too, I suggest that you supplement your genome with a more comprehensive database of phage”contigs” for more interpretable results Since we’re effectively working with only 3 bases in converted libraries, it’s important that the reference sequences represent all the abundant sources of input DNA. If not, you may see anomalous coverage where the aligner mis-places reads derived from other sources on your expected reference. There’s always a risk of this in complex samples… which is why i think it’s good to be suspicious of methylation calls in regions of unusually high coverage”.

2.Reran the pipeline using a combined sequences as reference (pop + pop mitochondrial genome + phage lambda + plasmid puc19c) instead of running pipeline with with sequence as reference sequences.

NEB tech advice: “It sounds like you’re running with just one genome at a time… I don’t recommend that, especially not with converted libraries due to the potential mis-mapping and confused methylation calls. Maybe it’s worth another try with the complete genome including chrM and the control genomes.”

3.Estimated conversion efficiency using chrM and phage lambda which can both be used as negative controls (unmethylated).

NEB tech advice: “You can get an estimate of APOBEC activity by looking at chrM, which is unmethylated in eukaryotes I know of. Unfortunately there is not really a built-in TET2+BGT/ methylated control (you can get an approximation if you know of some imprinted region).”

```{r Loading_data}
cpg_calls <- read.delim("/home/cgumbi/methylation/refcombined/ABLG-47625_CpG.bedGraph", header = FALSE, comment.char = "t")
chg_calls <- read.delim("/home/cgumbi/methylation/refcombined/ABLG-47625_CHG.bedGraph", header = FALSE, comment.char = "t")
chh_calls <- read.delim("/home/cgumbi/methylation/refcombined/ABLG-47625_CHH.bedGraph", header = FALSE, comment.char = "t")
# names on the file 
# col 1. The chromosome/contig/scaffold name
# col 2. The start coordinate
# col 3. The end coordinate
# col 4. The methylation   percentage rounded to an integer
# col 5. The number of alignments/pairs reporting methylated bases
# col 6. The number of alignments/pairs reporting unmethylated bases
# Rename the columns
colnames(cpg_calls) <- c("chrom", "start", "end", "percent_methylation", "methylated_reads", "unmethylated_reads")
colnames(chg_calls) <- c("chrom", "start", "end", "percent_methylation", "methylated_reads", "unmethylated_reads")
colnames(chh_calls) <- c("chrom", "start", "end", "percent_methylation", "methylated_reads", "unmethylated_reads")
```

# CHH and CHG are also good places to assess APOBEC activity.
Looking at look at the fraction of observed reads by context (which can be even with low read coverage) and Mbias plots produced by methyldackel are useful for this.

```{r summary_stats_by_context}
# Adding cytosine context column too each file
cpg_calls$context <- "CpG"
chg_calls$context <- "CHG"
chh_calls$context <- "CHH"

# Combining all cotext into one data frame ---
meth_all <- bind_rows(cpg_calls, chg_calls, chh_calls)
write.csv(meth_all,"/home/cgumbi/methylation/refcombined/Methylation_all_context.csv", row.names = F)
meth_all <- read.csv("/home/cgumbi/methylation/refcombined/Methylation_all_context.csv")

#Remove rows with missing values 
meth_all <- na.omit(meth_all)

# Summary statistics
summary_stats <- meth_all %>%
  group_by(context) %>%
  summarise(
    mean_meth = mean(percent_methylation, na.rm = TRUE),
    median_meth = median(percent_methylation, na.rm = TRUE),
    sd_meth = sd(percent_methylation, na.rm = TRUE),
    .groups = "drop")

summary_meth_df <- as.data.frame(summary_stats)
write.csv(summary_meth_df,"~/methylation/Methyl47625/tables/summary_methylation_stats.csv", row.names = F) 

#Boxplot of methylation across contexts
all_context <- ggplot(meth_all, aes(x = context, y = percent_methylation, fill = context)) +
  geom_boxplot(outlier.alpha = 0.3, outlier.size = 0.5) +
  labs(title = "Methylation levels across contexts",
       x = "Context",
       y = "Methylation (%)") +
  theme_ipsum() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

gg_check(all_context)
# Save polot with all cytosine context
ggsave("~/methylation/Methyl47625/figures/all_context_methylation_plot.png", plot = all_context, width = 10, height = 6, dpi = 300)
```

# Looking into the context vs chromosome 

```{r context_chrom}
chrom_context <- meth_all %>%
  group_by(chrom, context) %>%
  mutate(total_reads = methylated_reads + unmethylated_reads) %>%
  filter(total_reads == 2) %>%
  filter(chrom %in% c("OR482510.1", "plasmid_puc19c", "phage_lambda"))

# Create the barplot
contextvschrom <- ggplot(chrom_context, aes(x = context, y = percent_methylation, fill = chrom)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Methylation per Context by Chrom",
       x = "Context",
       y = "Methylation",
       fill = "Chrom") +
  theme_ipsum()
ggsave("~/methylation/Methyl47625/figures/contextvschrom_plot.png", plot = contextvschrom, width = 10, height = 6, dpi = 300)

```



# Editing CpG dataframe

```{r editing_cpg_df}
#Remove rows with missing values 
cpg_calls <- na.omit(cpg_calls)
# Remove "genome_hic_" and "_RagTag" from the 'chrom' column
cpg_calls$chrom <- gsub("genome_hic_", "", cpg_calls$chrom)
cpg_calls$chrom <- gsub("_RagTag", "", cpg_calls$chrom)
```

# Visualize CpG methylation by chromosome

```{r methylation_by_chom}
cpg_calls <- cpg_calls %>%
  filter(!is.na(chrom), !is.na(percent_methylation))

cpg_box <- ggplot(cpg_calls, aes(x = chrom, y = percent_methylation, fill = chrom)) +
  geom_boxplot(outlier.alpha = 0.3, outlier.size = 0.5) +
  theme_ipsum() +
  labs(title = "CpG methylation by scaffold",
       subtitle = "OR482510.1 = chrM, phage_lambda = unmethylated control, plasmid_puc19c = methylated control",
       x = "Scaffold", 
       y = "Percentage methylation (%)") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))   # rotate labels
gg_check(cpg_box)
# Save CgG plot
ggsave("~/methylation/Methyl47625/figures/CpG_methylation_plot.png", plot = cpg_box, width = 10, height = 6, dpi = 300)
```

# Sub-setting CpG methylation data chromosome

```{r Subset_data_chromosomes}
chrom_list <- split(cpg_calls, cpg_calls$chrom)
phage_lambda <- chrom_list[["phage_lambda"]] # unmethylated control
plasmid_puc19c <- chrom_list[["plasmid_puc19c"]] # methylated control 
chrM <-chrom_list[["OR482510.1"]] #POP mitochondrial genome
```

# Methylation level on methylated control pUC19

```{r pUC19}
# average methylation   on pUC19
puc19_df <- plasmid_puc19c %>%
   group_by(chrom) %>%
   summarize(avg_methylation = mean(percent_methylation, na.rm = TRUE),
     total_methylated_reads = sum(methylated_reads, na.rm = TRUE),
     total_unmethylated_reads = sum(unmethylated_reads, na.rm = TRUE),
     .groups = "drop")
write.csv(puc19_df,"~/methylation/Methyl47625/tables/puc19_methylation_summary.csv", row.names = F) 

# making a scatter plot
pUC19_scatter <- ggplot(plasmid_puc19c, aes(x = start, y = percent_methylation)) +
   geom_point(alpha = 1, size = 2) +
   labs(title = "Plasmid pUC19",
         subtitle = "Control DNA CpG methylated pUC19", 
         x = "Read position", y = "Percentage methylation (%)") +
    theme_ipsum()
gg_check(pUC19_scatter)

# frequency of methylation   percentage on pUC19
pUC19_hist <- ggplot(plasmid_puc19c, aes(x = percent_methylation)) +
   geom_histogram(binwidth = 5, fill = "#1f77b4", color = "white") +
   labs(title = "Methylation frequency of pUC19",
        subtitle = "Control DNA CpG methylated pUC19",
        x = "Percentage methylation (%)",y = "Frequency") + 
   theme_ipsum()
gg_check(pUC19_hist)

ggsave("~/methylation/Methyl47625/figures/pUC19_scatter_plot.png", plot = pUC19_scatter, width = 10, height = 6, dpi = 300)
ggsave("~/methylation/Methyl47625/figures/pUC19_hist_plot.png", plot = pUC19_hist, width = 10, height = 6, dpi = 300)
```

# Methylation on control (phage_lambda)

```{r lambda}
# average methylation on phage labmda
lambda_df <- phage_lambda %>%
   group_by(chrom) %>%
   summarize(
     avg_methylation   = mean(percent_methylation  , na.rm = TRUE),
     total_methylated_reads = sum(methylated_reads, na.rm = TRUE),
     total_unmethylated_reads = sum(unmethylated_reads, na.rm = TRUE),
     .groups = "drop")
write.csv(lambda_df,"~/methylation/Methyl47625/tables/lambda_methylation_summary.csv", row.names = F) 
 
# making a scatter plot
lambda_scatter <- ggplot(phage_lambda, aes(x = start, y = percent_methylation  )) +
   geom_point(alpha = 0.6, size = 2) +
   labs(title = "Lambda",
        subtitle = "Control DNA unmethylated lambda",
        x = "Read position",
        y = "Percentage methylation (%)") +
   theme_ipsum()
gg_check(lambda_scatter)

# Histogram of methylation   percentage
lambda_hist <- ggplot(phage_lambda, aes(x = percent_methylation  )) +
  geom_histogram(binwidth = 5, fill = "#2ca02c", color = "white") +
   labs(title = "Methylation frequency lambda",
        subtitle = "Control DNA unmethylated lambda",
        x = "Percentage methylation (%)", y = "Frequency") +
   theme_ipsum()
gg_check(lambda_hist)

ggsave("~/methylation/Methyl47625/figures/lambda_scatter_plot.png", plot = lambda_scatter, width = 10, height = 6, dpi = 300)
ggsave("~/methylation/Methyl47625/figures/lambda_hist.plot.png", plot = lambda_hist, width = 10, height = 6, dpi = 300)
```

# Conversion efficiency

The frequency of CpG methylation in the λ-bacteriophage genome gives an estimate of the conversion efficiency, as the λ-bacteriophage genome is grown in methylase-negative Escherichia coli, which cannot methylate cytosines in CpG context. Thus, for example, if the observed CpG methylation frequency in the λ-bacteriophage is 0.2%, the cytosine-to-thymine conversion in the sample is estimated to be 99.8% efficient (Singer 2019 : Apractical guide to the Measurement and analyses of DNA mmethylation).

```{r con_eff_on_labmda}
# conversion efficiency 
conv_eff_lambda <- 100 - mean(phage_lambda$percent_methylation  , na.rm = TRUE)
cat("APOBEC Conversion efficiency from phage Lambda:", round(conv_eff_lambda, 2), "%\n")
#APOBEC Conversion efficiency from phage Lambda: 96.55 %
```

# Methylation on the mitochondrail genome (chrM)

```{r chrM}
#average methylation on chrM
chrM_df <- chrM %>%
   group_by(chrom) %>%
   summarize(
     avg_methylation = mean(percent_methylation, na.rm = TRUE),
     total_methylated_reads = sum(methylated_reads, na.rm = TRUE),
     total_unmethylated_reads = sum(unmethylated_reads, na.rm = TRUE),
     .groups = "drop")

write.csv(chrM_df,"~/methylation/Methyl47625/tables/chrM_methylation_summary.csv", row.names = F) 

# making a scatter plot
chrM_scatter <- ggplot(chrM, aes(x = start, y = percent_methylation)) +
   geom_point(alpha = 0.6, size = 2) +
   labs(title = "CpG methylation on chrM",
        x = "Read position",
        y = "Percentage methylation (%)") +
   theme_ipsum()
gg_check(chrM_scatter)
 
# Histogram of methylation   percentage on chrM
chrM_hist <- ggplot(chrM, aes(x = percent_methylation  )) +
   geom_histogram(binwidth = 5, fill = "#d62728", color = "white") +
   labs(title = "Methylation frequency of chrM", 
        x = "Percentage methylation (%)",y = "Frequency") + 
   theme_ipsum()
gg_check(chrM_hist)

ggsave("~/methylation/Methyl47625/figures/chrM_scatter_plot.png", plot = chrM_scatter, width = 10, height = 6, dpi = 300)
ggsave("~/methylation/Methyl47625/figures/chrM_hist.plot.png", plot = chrM_hist, width = 10, height = 6, dpi = 300)

```

# Conversion efficiency on chrM

```{r con_eff_on_chrM}
## conversion efficiency chrM
conv_eff_chrM <- 100 - mean(chrM$percent_methylation  , na.rm = TRUE)
cat("APOBEC Conversion efficiency from chrM:", round(conv_eff_chrM, 2), "%\n")
#APOBEC Conversion efficiency from chrM: 96.91 %
```


# Methylation bias plots with ggplot

```{r mbias_plots}
puc19_mbias <- ggplot(plasmid_puc19c, aes(x = start, y = percent_methylation)) +
  geom_line() +
  labs(title = "Methylation bias plot for pUC19",
        subtitle = "Control DNA CpG methylated pUC19",
      x = "Read Position",
      y = "Percentage methylation (%)") +
    theme_ipsum()
gg_check(puc19_mbias)

lambda_mbias <- ggplot(phage_lambda, aes(x = start, y = percent_methylation)) +
  geom_line() +
  labs(title = "Methylation bias plot for Lambda",
        subtitle = "Control DNA unmethylated Lambda",
      x = "Read Position",
      y = "Percentage methylation (%)") +
    theme_ipsum()
gg_check(lambda_mbias)

chrM_mbias <- ggplot(chrM, aes(x = start, y = percent_methylation)) +
  geom_line() +
  labs(title = "Methylation bias plot for chrM",
        subtitle = "Control DNA mitochondrial genome",
      x = "Read Position",
      y = "Percentage methylation (%)") +
    theme_ipsum()
gg_check(chrM_mbias)

ggsave("~/methylation/Methyl47625/figures/chrM_mbias_plot.png", plot = chrM_mbias, width = 10, height = 6, dpi = 300)
ggsave("~/methylation/Methyl47625/figures/lambda_mbias.plot.png", plot = lambda_mbias, width = 10, height = 6, dpi = 300)
ggsave("~/methylation/Methyl47625/figures/puc19_mbias.plot.png", plot = puc19_mbias, width = 10, height = 6, dpi = 300)
```


# Looking at conversion efficiency on chg methylation 
NEB Tech advice on using CHH and CHG methylation to suplement controls amd chrM in tetcing for conversion effciency " If I wanted to judge the conversion efficiency I would simply look at the lowest methylation you see anywhere. Often it is enough to just look at the methylation levels in non-CG context (in mammalian systems), else some people look at reads from the MT, or withing CpG islands. If you see a general methylation of 0.3% in non-CG context, then your conversion efficiency must have been at least 99.7%"

```{r con_eff_on_chg }
#Remove rows with missing values 
chg_calls <- na.omit(chg_calls)
# methylation level summary 
chg_df <- chg_calls %>%
   group_by(chrom) %>%
   summarize(
     avg_methylation   = mean(percent_methylation  , na.rm = TRUE),
     total_methylated_reads = sum(methylated_reads, na.rm = TRUE),
     total_unmethylated_reads = sum(unmethylated_reads, na.rm = TRUE),
     .groups = "drop")

write.csv(chg_df,"~/methylation/Methyl47625/tables/chg_methylation_summary.csv", row.names = F) 

conv_eff_chg <- 100 - mean(chg_calls$percent_methylation  , na.rm = TRUE)
cat("APOBEC Conversion efficiency from chrM:", round(conv_eff_chg, 2), "%\n")

```

# Conversion efficiency with chh methylation 
Here I will apply similar logic as in the chg case above.  

```{r con_eff_on_chh}
#Remove rows with missing values 
chh_calls <- na.omit(chh_calls)
# methylation level summary
chh_df <- chh_calls %>%
   group_by(chrom) %>%
   summarize(
     avg_methylation   = mean(percent_methylation  , na.rm = TRUE),
     total_methylated_reads = sum(methylated_reads, na.rm = TRUE),
     total_unmethylated_reads = sum(unmethylated_reads, na.rm = TRUE),
     .groups = "drop")

write.csv(chh_df,"~/methylation/Methyl47625/tables/chh_methylation_summary.csv", row.names = F) 

conv_eff_chh <- 100 - mean(chh_calls$percent_methylation  , na.rm = TRUE)
cat("APOBEC Conversion efficiency from chrM:", round(conv_eff_chh, 2), "%\n")
```

# Table summary for conversion efficiency 

```{r sum_con_eff}
con_eff <- data.frame(approach = c("chrM","Lambda", "CHG methylation", "CHH methylation"),
                      conversion_efficiency = c(96.91, 96.55, 96.25,96.37 ))


con_eff_plot <- ggplot(con_eff, aes(x = approach, y = conversion_efficiency, fill = approach)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +  # Adjust bar width here
  labs(title = "APOBEC Conversion efficiency",
       subtitle = "CHG = 96.25%, CHH = 96.37%, chrM = 96.91%, Lambda = 96.55%", 
       x = "Approach",
       y = "Conversion efficiency (%)") +
  theme_ipsum() + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))  # rotate labels
gg_check(con_eff_plot)

ggsave("~/methylation/Methyl47625/figures/APOBEC_conversion_efficency_plot.png", plot = con_eff_plot, width = 10, height = 6, dpi = 300)
write.csv(con_eff,"~/methylation/Methyl47625/tables/APOBEC_conversion_efficiency.csv", row.names = F) 
```











