---
title: ABLG methylation analyses using all reference genome comined into one fasta
  file
author: "charles-gumbi"
date: "2025-08-13"
output:
  html_document:
    df_print: paged
---
```{r Loading_packages}
library(dplyr)
library(ggplot2)
library(readr)
library(hrbrthemes)
library(gcookbook)
library(extrafont)
library(tidyverse)
```

# Modifications from last analyses
1.Removed T4 and XP-12 phages from the in reference alignment

NEB tech support “We include T4 and XP-12 phages in our control references for hydroxymethylation assessment (a related product). Since you did not spike in T4 or XP-12 to your libraries you can likely ignore any reads that that happen to align to these contigs. If your fin samples do actually contain phages related to XP-12 and T4, you may see more coverage. If you’re interested in the phage genomics too, I suggest that you supplement your genome with a more comprehensive database of phage”contigs” for more interpretable results Since we’re effectively working with only 3 bases in converted libraries, it’s important that the reference sequences represent all the abundant sources of input DNA. If not, you may see anomalous coverage where the aligner mis-places reads derived from other sources on your expected reference. There’s always a risk of this in complex samples… which is why i think it’s good to be suspicious of methylation calls in regions of unusually high coverage”.

2.Reran the pipeline using a combined sequences as reference (pop + pop mitochondrial genome + phage lambda + plasmid puc19c) instead of running pipeline with with sequence as reference sequences.

NEB tech advice: “It sounds like you’re running with just one genome at a time… I don’t recommend that, especially not with converted libraries due to the potential mis-mapping and confused methylation calls. Maybe it’s worth another try with the complete genome including chrM and the control genomes.”

3.Estimated conversion efficiency using chrM and phage lambda which can both be used as negative controls (unmethylated).

NEB tech advice: “You can get an estimate of APOBEC activity by looking at chrM, which is unmethylated in eukaryotes I know of. Unfortunately there is not really a built-in TET2+BGT/ methylated control (you can get an approximation if you know of some imprinted region).”

```{r Loading_data}
cpg_calls <- read.delim("~/pop_methylation/Methyl47625/data/ABLG-47625_CpG.bedGraph.txt", header = FALSE, comment.char = "t")
#chg_calls <- read.delim("~/pop_methylation/Methyl47625/data/ABLG-47625_CHG.bedGraph.txt", header = FALSE, comment.char = "t")
#chh_calls <- read.delim("~/pop_methylation/Methyl47625/data/ABLG-47625_CHH.bedGraph.txt", header = FALSE, comment.char = "t")
# names on the file 
# col 1. The chromosome/contig/scaffold name
# col 2. The start coordinate
# col 3. The end coordinate
# col 4. The methylation   percentage rounded to an integer
# col 5. The number of alignments/pairs reporting methylated bases
# col 6. The number of alignments/pairs reporting unmethylated bases
# Rename the columns
colnames(cpg_calls) <- c("chrom", "start", "end", "percent_methylation", "methylated_reads", "unmethylated_reads")
#colnames(chg_calls) <- c("chrom", "start", "end", "percent_methylation", "methylated_reads", "unmethylated_reads")
#colnames(chh_calls) <- c("chrom", "start", "end", "percent_methylation", "methylated_reads", "unmethylated_reads")
```

# Sub-setting CpG methylation data chromosome

```{r Subset_data_chromosomes}
chrom_list <- split(cpg_calls, cpg_calls$chrom)
phage_lambda <- chrom_list[["phage_lambda"]] # unmethylated control
plasmid_puc19c <- chrom_list[["plasmid_puc19c"]] # methylated control 
chrM <-chrom_list[["OR482510.1"]] #POP mitochondrial genome
```


# Methylation level on methylated control pUC19

```{r pUC19}
# average methylation   on pUC19
puc19_df <- plasmid_puc19c %>%
   group_by(chrom) %>%
   summarize(avg_methylation = mean(percent_methylation, na.rm = TRUE),
     total_methylated_reads = sum(methylated_reads, na.rm = TRUE),
     total_unmethylated_reads = sum(unmethylated_reads, na.rm = TRUE),
     .groups = "drop")
puc19_df

# making a scatter plot
pUC19_scatter <- ggplot(plasmid_puc19c, aes(x = start, y = percent_methylation)) +
   geom_point(alpha = 1, size = 2) +
   labs(title = "Plasmid pUC19",
         subtitle = "Control DNA CpG methylated pUC19", 
         x = "Read position", y = "Percentage methylation (%)") +
    theme_ipsum()
pUC19_scatter
gg_check(pUC19_scatter)

# frequency of methylation   percentage on pUC19
pUC19_hist <- ggplot(plasmid_puc19c, aes(x = percent_methylation)) +
   geom_histogram(binwidth = 5, fill = "#1f77b4", color = "white") +
   labs(title = "Methylation frequency of pUC19",
        subtitle = "Control DNA CpG methylated pUC19",
        x = "Percentage methylation (%)",y = "Frequency") + 
   theme_ipsum()
pUC19_hist
gg_check(pUC19_hist)
```

# Methylation on control (phage_lambda)

```{r lambda}
# average methylation on phage labmda
lambda_df <- phage_lambda %>%
   group_by(chrom) %>%
   summarize(
     avg_methylation   = mean(percent_methylation  , na.rm = TRUE),
     total_methylated_reads = sum(methylated_reads, na.rm = TRUE),
     total_unmethylated_reads = sum(unmethylated_reads, na.rm = TRUE),
     .groups = "drop")
 lambda_df

# making a scatter plot
lambda_scatter <- ggplot(phage_lambda, aes(x = start, y = percent_methylation  )) +
   geom_point(alpha = 0.6, size = 2) +
   labs(title = "Lambda",
        subtitle = "Control DNA unmethylated lambda",
        x = "Read position",
        y = "Percentage methylation (%)") +
   theme_ipsum()
lambda_scatter
gg_check(lambda_scatter)

# Histogram of methylation   percentage
lambda_hist <- ggplot(phage_lambda, aes(x = percent_methylation  )) +
  geom_histogram(binwidth = 5, fill = "#2ca02c", color = "white") +
   labs(title = "Methylation frequency lambda",
        subtitle = "Control DNA unmethylated lambda",
        x = "Percentage methylation (%)", y = "Frequency") +
   theme_ipsum()
lambda_hist
gg_check(lambda_hist)
```

# Conversion efficiency

The frequency of CpG methylation in the λ-bacteriophage genome gives an estimate of the conversion efficiency, as the λ-bacteriophage genome is grown in methylase-negative Escherichia coli, which cannot methylate cytosines in CpG context. Thus, for example, if the observed CpG methylation frequency in the λ-bacteriophage is 0.2%, the cytosine-to-thymine conversion in the sample is estimated to be 99.8% efficient (Singer 2019 : Apractical guide to the Measurement and analyses of DNA mmethylation).

```{r Conversion_efficiency_on_labmda}
# conversion efficiency 
conv_eff_lambda <- 100 - mean(phage_lambda$percent_methylation  , na.rm = TRUE)
cat("APOBEC Conversion efficiency from phage Lambda:", round(conv_eff_lambda, 2), "%\n")
#APOBEC Conversion efficiency from phage Lambda: 96.55 %
```

# Methylation on the mitochondrail genome (chrM)

```{r chrM}
#average methylation on chrM
chrM_df <- chrM %>%
   group_by(chrom) %>%
   summarize(
     avg_methylation = mean(percent_methylation, na.rm = TRUE),
     total_methylated_reads = sum(methylated_reads, na.rm = TRUE),
     total_unmethylated_reads = sum(unmethylated_reads, na.rm = TRUE),
     .groups = "drop")
chrM_df
# making a scatter plot
chrM_scatter <- ggplot(chrM, aes(x = start, y = percent_methylation)) +
   geom_point(alpha = 0.6, size = 2) +
   labs(title = "CpG methylation on chrM",
        x = "Read position",
        y = "Percentage methylation (%)") +
   theme_ipsum()
chrM_scatter
gg_check(chrM_scatter)
 
# Histogram of methylation   percentage on chrM
chrM_hist <- ggplot(chrM, aes(x = percent_methylation  )) +
   geom_histogram(binwidth = 5, fill = "#d62728", color = "white") +
   labs(title = "Methylation frequency of chrM", 
        x = "Percentage methylation (%)",y = "Frequency") + 
   theme_ipsum()
chrM_hist
gg_check(chrM_hist)
```

# Conversion efficiency on chrM

```{r Conversion_efficiency_chrM}
## conversion efficiency chrM
conv_eff_chrM <- 100 - mean(chrM$percent_methylation  , na.rm = TRUE)
cat("APOBEC Conversion efficiency from chrM:", round(conv_eff_chrM, 2), "%\n")
#APOBEC Conversion efficiency from chrM: 96.91 %
```




